Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-04-20 marginal.Rmd
Revision of results in the elderly category

The general pattern and development of mortality resembles that from the younger age group, according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs). Median death age among women, as expected, was higher than among men. We find a clear increase in life expectancy, but
the fastest decline did not set in definitely until sometime after the
Second World War; that is, later than for the age
group 40--64 years. This corresponds to the Swedish development whereby we
find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.

The excess years of remaining life in different social classes compared to workers are shown in Figure \@ref(fig:cumsw2). Concerning the differences in the social pattern, it is less consisten compared to what
we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the same clear advantage for elite women during the period before 1950, but neither
do we find any large disadvantage for working-class elderly women at that
time. In most periods the social differences were small between working class women and those in the farming and middle categories (only differences of a couple of months in median age at death). The social differences continued to be small until the last decades of the 20th century. It is only during the four last periods (after the year 1975) that it is
possible to establish the expected and consistent social gradient in mortality.

For men we find indications of the same intriguing change among elderly as
for the adult male population aged 40--64 years. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
the 20th century. It is not until the second half of the 20th
century that the survival is better for the highest social class while
workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.

>>>>

The general pattern and development of mortality resembles that from the younger age group (notice that the comparisons are made with different methods in the two age groups, and the figures are therefore only comparable for the identification of patterns in the development), according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs). Median death age among women, as expected, was higher than among men. We find a clear increase in life expectancy, but
the fastest decline did not set in definitely until sometime after the
Second World War; that is, later than for the age
group 40--64 years. This corresponds to the Swedish development whereby we
find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.

The excess years of remaining life in different social classes compared to workers are shown in Figure \@ref(fig:cumsw2). Concerning the differences in the social pattern, it is less consistent compared to what
we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the same clear advantage for elite women during the period before 1925, but neither
do we find any large disadvantage for working-class elderly women at that
time. In most periods the social differences were small between working class women and those in the farming and middle categories (only differences of a couple of months in median age at death). The social differences continued to be small until the last decades of the 20th century. It is only during the four last periods (after the year 1975) that it is
possible to establish the expected and consistent social gradient in mortality.

For men we find indications of the same intriguing change among elderly as
for the adult male population aged 40--64 years, but with a more mixed social pattern in this age group until the last periods. There was no clear social gradient during the 19th century, a pattern that prevailed until the middle of the the 20th century. Similar to what we found for men in working age, it is not until the second half of the 20th
century that a distinct social gradient appears, characterised by better survival for higher social classes. The difference between the worker and the farmer group, that towards the end of the 20th century has become marginal in numbers has been small in most periods.
